Synthesizing Quotient Lenses

### 2017-10-03 papers/QuotientLensSynth/main.tex
added a table for the K and W functions

\newcommand{\FINISH}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\bcp}[1]{\FINISH{dkred}{B}{#1}}
\newcommand{\BCP}[1]{\FINISH{dkred}{B}{\bf #1}}
\newcommand{\afm}[1]{\FINISH{dkgreen}{A}{#1}}
\newcommand{\dpw}[1]{\FINISH{dkblue}{D}{#1}}
\newcommand{\saz}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\ksf}[1]{\FINISH{teal}{K}{#1}}
\newcommand{\sm}[1]{\FINISH{dkpurple}{SM}{#1}}

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathit{#1}}}

@@

$$perm(\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER}, \re{YEAR}) \; with \; (\re{WSP}^* \mapsto \string
\n),$$

@@

\sm{[Mention and give examples of quotient lenses here?]}

@@

  \paragraph{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \; | \; q \mapsto s \; | \; squash (R, R', f) \; | \;
perm \; (q_1, \ldots, q_n) \; with \; q \; | \; normalize(R, R', f)\\
&q' \circ q \; | \; q \cdot q' \; | \; (q \; | \; q') \; | \; q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions whose
domain and codomain are regular languages, and $s$ ranges over character
strings.

@@

We say that a regular expression $R$ is \textit{strongly unambiguous} if and
only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1, S_2$
strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \; | \; S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

We now give the semantics of QREs as well as inference rules for deriving them:
\textcolor{red}{[Informal Meanings/Example Usages Needed]}\\
\begin{enumerate}
  \item $q = R$
\begin{center}
\begin{enumerate}
  \item $W(R) = R$,
  \item $K(R) = R$, 
  \item
For all $w, w' \in W(R)$, $w \; EqRel(q) \; w'$ if and only if $w =
w'$.
\item
$Canonizer(R)(r) = r$
\end{enumerate}
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}
\end{center}
\item $q=q \mapsto s$
\begin{center}
\begin{enumerate}
  \item $W(q \mapsto s) = W(q)$,
  \item $K(q \mapsto s) = s$, 
  \item
For all $w, w' \in W(q \mapsto S)$, $w \; EqRel(q \mapsto S) \; w'$,
\item
$Canonizer(q \mapsto S)(r) = s$
\end{enumerate}
\begin{prooftree}
\AxiomC{$s \in W(q)$}
\UnaryInfC{$q \mapsto s$ is well formed}
\end{prooftree}
\end{center}
\item $q=squash(R, R', f)$
\begin{center}
\begin{enumerate}
  \item $W(squash(R, R', f)) = R \; | \; R'$,
  \item $K(squash(R, R', f)) = R'$, 
  \item
For all $w, w' \in W(squash(R, R', f))$, $w \; EqRel(q) \; w'$ if and only
if $f(w) = w'$ or $w = w'$
\item
$Canonizer(squash(R, R', f))(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}$
\end{enumerate}
\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', f)$ is well formed}
\end{prooftree}
\end{center}
\item $q=perm \; (q_1, \ldots, q_n) \; with \; q$. Given regular expressions
$R_1, \ldots, R_n$ and $S$, let $\prod({R_i}, S) = R_1 \cdot S \cdot \ldots
\cdot S \cdot R_n$.
\begin{center}
\begin{enumerate}
  \item $W(perm \; (q_1, \ldots, q_n) \; with \; q) = \bigcup \limits_{\sigma
  \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot
  W(q_{\sigma(n)}),$
  \item $K(perm \; (q_1, \ldots, q_n) \; with \; q) = K(q_1) \cdot K(q_1)
  \cdot \ldots \cdot \cdot K(q) \cdot K(q_n)$, 
  \item
For all $w, w' \in W(q)$,  $w \; EqRel(q) \; w'$ if and only if
   $$w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot
    r_{\sigma(n)} \text{ and } r' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots
    \cdot s'_{n-1} \cdot {r'}_{\theta(n)}$$ for some $\sigma, \theta \in
    S_n$, with $r_i \; EqRel(q_i) \; r'_i$ and $s_k \; EqRel(q) \; s'_{k}$.
\item
$Canonize(q)(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonize(q_1)(r_1)) \cdot (Canonize(q)(s_1)) \cdot \ldots \cdot \cdot \cdot
(Canonize(q_n)(r_n))$
\end{enumerate}
\begin{prooftree}
\AxiomC{$\forall \sigma \neq \theta, \; \prod(\{W(q_{\sigma(i)})\}, W(q))\cap
 \prod(\{W(q_{\theta(i)})\}, W(q)) = \varnothing$}
\AxiomC{$\forall \sigma, \; K(q_{\sigma(1)}) \cdot^! K(q) \cdot^! \ldots 
\cdot^! K(q_{\sigma(n)})$}
\BinaryInfC{$permute \; (q_1, \ldots, q_n) \; with \;  q$ is well formed}
\end{prooftree}
\end{center}

@@

\item $q=q \; | \;  q'$

@@

  \item $W(q \cdot q') = W(q) \; | \; W(q')$,
  \item $K(q \cdot q') = K(q) \; | \; K(q')$, 

@@

For all $w, w' \in W(q \; | \; q')$, $w \; EqRel(q' \; | \; q) \; w'$ if and

@@

$Canonizer(q \; | \; q') = Canonizer(q) \; | \; Canonizer(q')

@@

$$\ell := R \; | \; const(s, s') \; | \;  swap(\ell,
\ell) \; | \; \ell \cdot \ell' \; |  \; (\ell \; | \; \ell') \; | \; \ell^* \;

@@

\llbracket R \rrbracket &= \{(r, r) \; | \; r \in \mathcal{L}(R)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \; | \;

@@

\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \; | \;

@@

\llbracket \ell \; | \; \ell' \rrbracket &= \{(s \cdot t) \; | \;

@@

\cdot t_n) \; | \; (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1

@@

$$ q := id(\ell) \; | \; lquot(q, \ell) \; | \; rquot(\ell, q) \; | \; q
\cdot q_2 \; | \; (q \; | \; q') \; | \; q^* \; | \; q \circ q',$$

@@

  \item
  $q = q_1 \; | \; q_2$, then
      \begin{prooftree}
\AxiomC{$q_1 : c_1 \Leftrightarrow d_1 $}
\AxiomC{$q_2 : c_2 \Leftrightarrow d_2$}
\AxiomC{$\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing$}
\AxiomC{$\mathcal{L}(W(d_1)) \cap \mathcal{L}(W(d_2)) = \varnothing$}
\QuaternaryInfC{$q_1 \; | \; q_2: (c_1 \; | \; c_2)
\Leftrightarrow (d_1 \; | \; d_2)$}
\end{prooftree}
  $$
  \llbracket q_1 \; | \; q_2 \rrbracket.get(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_1))\\
  \llbracket q_2 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_2))\\
  \end{cases}$$
  $$\llbracket q_1 \; | \; q_2 \rrbracket.put(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_1))\\
  \llbracket q_2 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_2))\\
  \end{cases}
  $$
\end{enumerate}

@@

  $q = q_1 \; | \; q_2$ where $q_1 : c_1 \Leftrightarrow d_1 $, $q_2 : c_2

@@

  \llbracket q_1 \; | \; q_2 \rrbracket.get(s) = 

@@

  $$\llbracket q_1 \; | \; q_2 \rrbracket.put(s) = 

@@

  so $\llbracket q_1 \; | \; q_2 \rrbracket.get = \llbracket \ell_1 \; | \;
  \ell_2 \rrbracket \circ Canonizer(c_1 \; | \; c_2)$. A similar argument shows
  that $\llbracket q_1 \; | \; q_2 \rrbracket.put = \llbracket \ell_1 \; | \;
  \ell_2 \rrbracket^{-1} \circ Canonizer(d_1 \; | \; d_2)$.\\

>>>>

\newcommand{\FINISH}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\bcp}[1]{\FINISH{dkred}{B}{#1}}
\newcommand{\BCP}[1]{\FINISH{dkred}{B}{\bf #1}}
\newcommand{\afm}[1]{\FINISH{dkgreen}{A}{#1}}
\newcommand{\dpw}[1]{\FINISH{dkblue}{D}{#1}}
\newcommand{\saz}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\ksf}[1]{\FINISH{teal}{K}{#1}}
\newcommand{\sam}[1]{\FINISH{dkpurple}{SM}{#1}}

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash}(#1, #2, #3)}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}(#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize}(#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}

@@

$$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER}, \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)},$$

@@

\sam{[Mention and give examples of quotient lenses here?]}

@@

  \paragraph{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep q \mapsto s \sep \squash{R}{R'}{f} \sep
\perm{(q_1, \ldots, q_n)}{q} \;  | \; \normalize{R}{R'}{f}\\
&q' \circ q \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions whose
domain and codomain are regular languages, and $s$ ranges over character
strings.

@@

We say that a regular expression $R$ is \textit{strongly unambiguous} if and
only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1, S_2$
strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

We now give the semantics of QREs as well as inference rules for deriving them:
\textcolor{red}{[Informal Meanings/Example Usages Needed]}

\begin{figure}[t]
  \centering
  \[
    \begin{array}{r@{\quad}c@{\quad}cc}

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      q \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q_1^* & W(q_1)^* & K(q_1)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q_1) \cdot \ldots \cdot \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions.}
  \label{fig:wk}
\end{figure}

\begin{enumerate}
  \item $q = R$
\begin{center}
\begin{enumerate}
  \item $W(R) = R$,
  \item $K(R) = R$, 
  \item
For all $w, w' \in W(R)$, $w \; EqRel(q) \; w'$ if and only if $w =
w'$.
\item
$Canonizer(R)(r) = r$
\end{enumerate}
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}
\end{center}
\item $q=q \mapsto s$
\begin{center}
\begin{enumerate}
  \item $W(q \mapsto s) = W(q)$,
  \item $K(q \mapsto s) = s$, 
  \item
For all $w, w' \in W(q \mapsto S)$, $w \; EqRel(q \mapsto S) \; w'$,
\item
$Canonizer(q \mapsto S)(r) = s$
\end{enumerate}
\begin{prooftree}
\AxiomC{$s \in W(q)$}
\UnaryInfC{$q \mapsto s$ is well formed}
\end{prooftree}
\end{center}
\item $q=squash(R, R', f)$
\begin{center}
\begin{enumerate}
  \item $W(squash(R, R', f)) = R \sep R'$,
  \item $K(squash(R, R', f)) = R'$, 
  \item
For all $w, w' \in W(squash(R, R', f))$, $w \; EqRel(q) \; w'$ if and only
if $f(w) = w'$ or $w = w'$
\item
$Canonizer(squash(R, R', f))(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}$
\end{enumerate}
\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', f)$ is well formed}
\end{prooftree}
\end{center}
\item $q=perm \; (q_1, \ldots, q_n) \; with \; q$. Given regular expressions
$R_1, \ldots, R_n$ and $S$, let $\prod({R_i}, S) = R_1 \cdot S \cdot \ldots
\cdot S \cdot R_n$.
\begin{center}
\begin{enumerate}
  \item $W(perm \; (q_1, \ldots, q_n) \; with \; q) = \bigcup \limits_{\sigma
  \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot
  W(q_{\sigma(n)}),$
  \item $K(perm \; (q_1, \ldots, q_n) \; with \; q) = K(q_1) \cdot K(q_1)
  \cdot \ldots \cdot \cdot K(q) \cdot K(q_n)$, 
  \item
For all $w, w' \in W(q)$,  $w \; EqRel(q) \; w'$ if and only if
   $$w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot
    r_{\sigma(n)} \text{ and } r' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots
    \cdot s'_{n-1} \cdot {r'}_{\theta(n)}$$ for some $\sigma, \theta \in
    S_n$, with $r_i \; EqRel(q_i) \; r'_i$ and $s_k \; EqRel(q) \; s'_{k}$.
\item
$Canonize(q)(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonize(q_1)(r_1)) \cdot (Canonize(q)(s_1)) \cdot \ldots \cdot \cdot \cdot
(Canonize(q_n)(r_n))$
\end{enumerate}
\begin{prooftree}
\AxiomC{$\forall \sigma \neq \theta, \; \prod(\{W(q_{\sigma(i)})\}, W(q))\cap
 \prod(\{W(q_{\theta(i)})\}, W(q)) = \varnothing$}
\AxiomC{$\forall \sigma, \; K(q_{\sigma(1)}) \cdot^! K(q) \cdot^! \ldots 
\cdot^! K(q_{\sigma(n)})$}
\BinaryInfC{$permute \; (q_1, \ldots, q_n) \; with \;  q$ is well formed}
\end{prooftree}
\end{center}

@@

\item $q=q \sep  q'$

@@

  \item $W(q \cdot q') = W(q) \sep W(q')$,
  \item $K(q \cdot q') = K(q) \sep K(q')$, 

@@

For all $w, w' \in W(q \sep q')$, $w \; EqRel(q' \sep q) \; w'$ if and

@@

$Canonizer(q \sep q') = Canonizer(q) \sep Canonizer(q')

@@

$$\ell := R \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;

@@

\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep

@@

\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep

@@

\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep

@@

\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1

@@

$$ q := id(\ell) \sep lquot(q, \ell) \sep rquot(\ell, q) \sep q
\cdot q_2 \sep (q \sep q') \sep q^* \sep q \circ q',$$

@@

  \item
  $q = q_1 \sep q_2$, then
      \begin{prooftree}
\AxiomC{$q_1 : c_1 \Leftrightarrow d_1 $}
\AxiomC{$q_2 : c_2 \Leftrightarrow d_2$}
\AxiomC{$\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing$}
\AxiomC{$\mathcal{L}(W(d_1)) \cap \mathcal{L}(W(d_2)) = \varnothing$}
\QuaternaryInfC{$q_1 \sep q_2: (c_1 \sep c_2)
\Leftrightarrow (d_1 \sep d_2)$}
\end{prooftree}
  $$
  \llbracket q_1 \sep q_2 \rrbracket.get(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_1))\\
  \llbracket q_2 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_2))\\
  \end{cases}$$
  $$\llbracket q_1 \sep q_2 \rrbracket.put(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_1))\\
  \llbracket q_2 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_2))\\
  \end{cases}
  $$
\end{enumerate}

@@

  $q = q_1 \sep q_2$ where $q_1 : c_1 \Leftrightarrow d_1 $, $q_2 : c_2

@@

  \llbracket q_1 \sep q_2 \rrbracket.get(s) = 

@@

  $$\llbracket q_1 \sep q_2 \rrbracket.put(s) = 

@@

  so $\llbracket q_1 \sep q_2 \rrbracket.get = \llbracket \ell_1 \sep
  \ell_2 \rrbracket \circ Canonizer(c_1 \sep c_2)$. A similar argument shows
  that $\llbracket q_1 \sep q_2 \rrbracket.put = \llbracket \ell_1 \sep
  \ell_2 \rrbracket^{-1} \circ Canonizer(d_1 \sep d_2)$.\\
